Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-03-25 doc/invt.tex
implement reflective boundary conditions for MD
invt.tex: edit abstract and introduction

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques
  in free energy calculations.
  %
  Both methods sample a flat distribution
  along a quantity of interest, $z$,
  by constructing a bias potential that offsets
  the potential of mean force (PMF) on the fly.
  %
  The two, however, differ by
  the manner of updating the bias potential:
  %
  in the WL case,
  each update applies to only
  the bin containing the current $z$,
  while in the metadynamics case,
  a Gaussian wave packet
  involving several neighboring bins
  is used.
  %
  The asymptotic error of the resulting PMF
  is determined by the updating schedule,
  or the updating magnitude
  as a function of simulation time.
  %
  In the WL case,
  the optimal schedule
  is known to be given by
  the inverse simulation time.
  %
  In this study, 
  we give a formula of computing the optimal schedule
  for a general updating scheme
  involving multiple bins.
  %
  We show that while metadynamics can be advantageous
  for relatively short simulations,
  the WL algorithm is best for asymptotic convergences.
  %the optimal schedule
  %and the effectiveness of metadynamics
  %are sensitive to the simulation length and other parameters.
  %
  %Further, we compared the two algorithms, and found that
  %although metadynamics produces a smoother PMF profile,
  %it is less effective in reducing errors with the shorter wavelengths.
  %%
  %Thus, the latter relies more heavily on the assumption of
  %the smoothness of the PMF profile,
  %especially with a wide updating window.
\end{abstract}

@@

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
the problem is to compute,
via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
and molecular dynamics\cite{frenkel, karplus2002} (MD) simulations,
a distribution, $p^*(z)$,
along a quantity of interest $z$.
%
The negative logarithm,
$-\log p^*(z)$,
defines a free energy,
or the potential of mean force (PMF).
%
Usually, the distribution $p^*(z)$ is
localized in one or a few small $z$ regions.
%
Therefore, to capture the global shape
of the PMF,
one has to run multiple independent simulations
under different umbrella or bias potentials
along $z$\cite{mezei1987, berg1992, lee1993},
and then piece together the information.
%
This solution is clumsy in practice
and often suffers from a convergence problem
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum
for a long time.

A better solution is to artificially alter
the target distribution along $z$
to a flat one\cite{mezei1987, berg1992, lee1993,
wang2001, wang2001pre, laio2002, laio2008, barducci2011, sutto2012}.
%
This alleviates the above problem of local trapping
by forcing the system to travel more 
deliberately along $z$,
which often represents
a direction of slow motion.
%
To achieve a flat distribution, however,
the method must be able to construct a
bias potential that exactly cancels the PMF.

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and meta\-dy\-nam\-ics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
techniques for this purpose,
in which a bias potential is actively built-up on-the-fly
during simulation.
%
These techniques employ updating loops
that discourage the system
to sample configurations
sharing the current $z$ value
by elevating the bias potential here.
%
The main difference of the two
is that in the WL case,
each updating step is limited to the bin
containing the current $z$,
while metadynamics adopts an extended
updating wave packet
covering several neighboring bins.
%
The latter is more often used in MD simulations
as it guarantees a smooth profile
of the bias potential
suitable for numerical differentiation
in deriving the bias force.
%
However, in both cases, the bias potential
is updated frequently (usually every few MC or MD steps),
which brings about a side effect 
of disrupting the underlying
equilibrium sampling dynamics\cite{laio2002}.
%
Thus, one often has to reduce
the updating magnitude
over the course of simulation
to decrease the error.

The updating schedule,
or the manner of reducing
the updating magnitude\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014},
therefore,
determines the precision of the final bias potential,
hence that of the PMF.
%
For the WL algorithm, it is known
that the optimal magnitude of updating $\ln f$
is given by the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008},
where ``time'' means
the number of simulation steps
divided by the number of bins.
%
The optimal schedule for metadynamics,
however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for a general updating scheme,
including those used in WL and metadynamics simulations.
%
Although the method recovers the inverse-time formula
for the WL case,
it predicts, for metadynamics,
a more algebraically complex schedule,
which is sensitive to the simulation length
and the width of the Gaussian wave packet.
%
Further, by comparing
different updating schemes,
we show that
the single-bin updating scheme
used by the WL algorithm
is best for infinitely long simulations.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of
flat-distribution sampling
and the development on the optimal schedule.
%
This also helps fixing the notations.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
on the simplest case of
single-bin updating scheme.
%
We shall prove the optimality
of the known inverse-time formula.
%
Next, we turn to the general case
of multiple-bin updating schemes
in Secs. \ref{sec:multiple-bin}
to \ref{sec:band-matrix}.
%
Finally, we compare different updating schemes
in the long time limit
in Sec. \ref{sec:cmpschemes}.

@@

Given a system,
consider the problem of computing
the distribution, $p^*_i$
along a discrete quantity $i$.
%
%The discreteness of $i$ is natural
%in discrete models.
%
For example, $i$ can be the energy $E$
in lattice spin models; or the temperature index
in a simulated tempering simulation.
%
For a continuous quantity, $z$,
which is typical in molecular systems,
we can discretize $z$
such that the integer $i$ represents
the index of a small interval, or a bin,
$(z, z + dz)$.
%
In both cases,
the distribution is normalized as
$\sum_{i = 1}^n p^*_i = 1$.

@@

Here we shall call a simulation that targets
a nearly flat distribution\cite{dayal2004, trebst2004, singh2011},
$p_i$,
a flat-distribution sampling (FDS) simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS % (E-FDS) methods hereinafter,
estimates the bias potential, $V_i$,
and fixes it during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{berg1992, lee1993}
and simulated tempering\cite{marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potential, however,
can often be improved in the end of simulation.
%
To do so,
we accumulate a histogram along $i$
during simulation and use it for $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}.
%
This allows us to correct $V_i$ as
%
\begin{equation}
V_i^{\mathrm{corr}}
=
V_i^{\mathrm{}}
+
\log \frac{ \pi_i } { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
For an unknown system,
the main difficulty of the process
is that it is hard to guess a sufficiently accurate
$V_i$ to start with.
%
Thus, 
it usually takes several iterative rounds of simulations
to achieve the desired flat distribution.
%
Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.

For this reason, the second type of methods,
referred to as the adaptive FDS methods,
were invented.
%
Examples include
the WL and metadynamics methods.
%
These methods update the bias potential more frequently
and are more capable of achieving a flat distribution.
%
However, they are less well understood theoretically,
the updating loops used there
break the microscopic reversibility
of the underlying sampling mechanism.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods gradually
reduce the updating magnitude
over the course of simulation\cite{marsili2006,
liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008}.
%
Thus, in late stages of a long simulation,
the updating is negligible and
such an adaptive simulation is rather similar
to the corresponding equilibrium simulation.
%
If the bias potential were converged to the exact value,
the equilibrium simulation can be thought as
the optimal asymptotic limit of the adaptive counterpart,
delivering the maximal possible efficiency.

For later convenience, we introduce a shifted bias potential
%
\begin{equation}
  v_i \equiv V_i - \ln p^*_i + \ln p_i.
  \label{eq:v_def}
\end{equation}
%
Since the last two terms of Eq. \eqref{eq:v_def}
remain constant during the course of simulation,
updates to $V_i$'s are equivalent to those to $v_i$'s.
%
In terms of $v_i$'s, Eq. \eqref{eq:pi_p_phi}
becomes
%
\begin{equation}
  \pi_i
  =
  \frac{                p_i \, e^{-v_i} }
       { \sum_{j = 1}^n p_j \, e^{-v_j} }
  \propto
  p_i \, e^{-v_i}.
  \label{eq:pi_p_v}
\end{equation}
%
According to Eq. \eqref{eq:pi_p_phi1},
$v_i$ should approach a constant of $i$
upon convergence (i.e., when $\pi_i \approx p_i$).

>>>>

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques
  in free energy calculations.
  %
  Both methods sample a flat distribution
  along a quantity of interest, $z$,
  by constructing a bias potential that offsets
  the potential of mean force (PMF) on the fly.
  %
  The two, however, differ by
  the manner of updating the bias potential:
  %
  the WL algorithm updates only
  the bin containing the current $z$,
  while metadynamics updates also
  several neighboring bins according to
  a Gaussian wave packet.
  %
  The error of the resulting PMF
  is determined by the updating schedule,
  or the updating magnitude
  as a function of simulation time.
  %
  In the WL case,
  the optimal schedule
  is known to be given by
  the inverse simulation time.
  %
  In this study,
  we give a formula of computing the optimal schedule
  for a general updating scheme
  involving multiple bins.
  %
  We show that while metadynamics can be advantageous
  for relatively short simulations,
  the WL algorithm is best for asymptotic convergence.
  %the optimal schedule
  %and the effectiveness of metadynamics
  %are sensitive to the simulation length and other parameters.
  %
  %Further, we compared the two algorithms, and found that
  %although metadynamics produces a smoother PMF profile,
  %it is less effective in reducing errors with the shorter wavelengths.
  %%
  %Thus, the latter relies more heavily on the assumption of
  %the smoothness of the PMF profile,
  %especially with a wide updating window.
\end{abstract}

@@

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
the problem is to compute,
via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
and molecular dynamics\cite{frenkel, karplus2002} (MD) simulations,
a distribution, $p^*(z)$,
along a quantity of interest $z$.
%
The negative logarithm,
$-\log p^*(z)$,
defines a free energy,
or the potential of mean force (PMF).
%
Normally, the distribution $p^*(z)$ is
limited in one or a few narrow $z$ regions.
%
Therefore, to capture the global shape
of the PMF,
one has to run multiple independent simulations
under different umbrella or bias potentials
along $z$\cite{mezei1987, berg1992, lee1993},
and then piece together the entire curve.
%
This solution is clumsy in practice
and often fails to converge quickly
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum
for a long time.

A better solution is to artificially flatten
the target distribution along $z$\cite{mezei1987, berg1992, lee1993,
wang2001, wang2001pre, laio2002, laio2008, barducci2011, sutto2012}.
%
This alleviates the above problem of local trapping
by forcing the system to travel more
deliberately along $z$,
which often represents
a direction of slow motion.
%
To achieve a flat distribution, however,
the method must be able to construct a
bias potential that exactly cancels the PMF.

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and meta\-dy\-nam\-ics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
techniques for this purpose,
in which a bias potential is actively built up on-the-fly
during simulation.
%
These techniques employ updating loops
that discourage future visits to configurations
with the current $z$ value
by elevating the bias potential here.
%
The main difference between the two
is that in the WL case,
each updating step is limited to the bin
containing the current $z$,
while metadynamics adopts an extended
Gaussian wave packet
covering several neighboring bins.
%
The latter is more often used in MD simulations
as it guarantees a smooth profile
of the bias potential
suitable for numerical differentiation
in deriving the bias force.
%
However, in both cases, the bias potential
is updated frequently (usually every few MC or MD steps),
which brings about a side effect
of disrupting the underlying
equilibrium sampling dynamics\cite{laio2002}.
%
Thus, one often has to reduce
the updating magnitude
over the course of simulation
to decrease the error.

The updating schedule,
or the manner of reducing
the updating magnitude\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014},
therefore,
determines the precision of the final bias potential,
hence that of the PMF.
%
For the WL algorithm, it is known
that the optimal magnitude of updating $\ln f$
is given by the product of the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008}
(with ``time'' being
the number of simulation steps)
and the number of bins.
%
The optimal schedule for the metadynamics case,
however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for a general updating scheme,
including those used in WL and metadynamics simulations.
%
The method reproduces the inverse-time formula
for the WL case.
%
It also predicts, for metadynamics,
a more algebraically complex schedule,
which is sensitive to the simulation length
and the width of the Gaussian wave packet.
%
Further, by comparing
different updating schemes,
we show that
the single-bin updating scheme
used by the WL algorithm
is best for infinitely long simulations.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of
flat-distribution sampling
and some known aspects on the optimal schedule.
%
This also helps fixing the notations.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
on the simplest case of
single-bin (WL) updating scheme.
%
We shall prove the optimality
of the known inverse-time formula.
%
Next, we turn to the general case
of multiple-bin updating schemes
in Secs. \ref{sec:multiple-bin}
to \ref{sec:band-matrix}.
%
Finally, we compare different updating schemes
in the long time limit
in Sec. \ref{sec:cmpschemes}.

@@

Given a system,
consider the problem of computing
the distribution, $p^*_i$,
along a discrete quantity $i$.
%
%The discreteness of $i$ is natural
%in discrete models.
%
For example, $i$ can be the energy $E$
in a lattice spin model; or the temperature index
in a simulated tempering simulation.
%
For a continuous quantity, $z$,
which is typical in molecular systems,
we can discretize $z$
such that the integer $i$ represents
the index of a small interval, or a bin,
$(z, z + dz)$.
%
In both cases,
the distribution is normalized as
$\sum_{i = 1}^n p^*_i = 1$.

@@

Here we shall call a simulation that targets
a nearly flat distribution\cite{dayal2004, trebst2004, singh2011},
$p_i$,
a flat-distribution sampling (FDS) simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
estimates the bias potential, $V_i$,
before simulation,
and fixes it during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{berg1992, lee1993}
and simulated tempering\cite{marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potential, $V_i$,
can often be improved
by the histogram along $i$ accumulated
during simulation
Using the histogram as $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\log \frac{ \pi_i }
          {   p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    \pi_i &\propto p^*_i \, e^{-V_i}, \\
    p_i &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
the main difficulty of the process
is that it is hard to guess a sufficiently accurate
$V_i$ to start with.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.

For this reason, the second type of methods,
referred to as the adaptive FDS methods,
were invented.
%
Examples include
the WL and metadynamics methods.
%
These methods update the bias potential more frequently
and can easily achieve a flat distribution.
%
However, they are less well understood theoretically,
as the updating loops used there
break the microscopic reversibility
of the underlying sampling mechanism.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods gradually
reduce the updating magnitude
over the course of simulation\cite{marsili2006,
liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008}.
%
Thus, in late stages of a long simulation,
the updating is negligible and
such an adaptive simulation is rather similar
to the corresponding equilibrium simulation.
%
If the bias potential were converged to the exact value,
this particular equilibrium simulation can be thought as
the optimal asymptotic limit of the adaptive counterpart,
delivering the maximal efficiency.

For later convenience, we introduce a shifted bias potential
%
\begin{equation}
  v_i \equiv V_i - \ln p^*_i + \ln p_i.
  \label{eq:v_def}
\end{equation}
%
Since the last two terms of Eq. \eqref{eq:v_def}
remain constant during the course of simulation,
updates to $V_i$'s are equivalent to those to $v_i$'s.
%
In terms of $v_i$'s, Eq. \eqref{eq:pi_p_phi}
becomes
%
\begin{equation}
  \pi_i
  =
  \frac{                p_i \, e^{-v_i} }
       { \sum_{j = 1}^n p_j \, e^{-v_j} }
  \propto
  p_i \, e^{-v_i}.
  \label{eq:pi_p_v}
\end{equation}
%
According to Eq. \eqref{eq:pi_p_phi1},
$v_i$ should approach a constant of $i$
upon convergence (i.e., when $\pi_i \approx p_i$)
of an adaptive FDS.
